---
title: "Floristic Quality Index"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
header-includes:
- \usepackage[width=\textwidth]{caption}
- \usepackage{enumitem}
- \usepackage{wrapfig}
always_allow_html: yes
csl: ../citations/citations/apa.csl
bibliography: ../citations/citations/citations.bib
link-citations: yes
---

\pagenumbering{gobble}
\vspace{-1cm}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(dpi = 300) 
knitr::opts_chunk$set(message = F)
knitr::opts_chunk$set(warning = F)
knitr::opts_chunk$set(fig.align="left") 
```

Floristic Quality Assessments (FQA) utilize the vascular plant species at a site as an indicator of habitat quality. The fundamental assumption guiding the use of plants as indicators of habitat quality is that different species respond differently to the types and frequencies of disturbance. At one end of this disturbance spectrum are species which are able to persist, or may be introduced to an area, after certain types of anthropogenic disturbance - e.g. compaction of soil via heavy vehicles. On the other end of the spectrum are taxa which may only grow in areas which receive episodic natural disturbances characteristic of their ecosystem - e.g. a 100 year flood in a wetland. The subjectively estimated likelihood of a species either persisting, or being removed from a site due to disturbance is expressed as a Conservatism Value (C-Value). C-Values range from 1 to 10 for plants native to North America, and 0 for plants introduced to the continent since colonization by European Settlers, with plants which are not tolerant of disturbance being at the upper end of the spectrum. 

The use of FQA are uncommon in the Bureau of Land Management, perhaps in part due to the FQA originating in the Midwest, and the assignment of C-values being a task which requires considerable amounts of resources (@spyreas2019floristic). A further requirement which hampers the utilization of these metrics are that each individual plant species, often times including subspecies, is assigned a separate C-value,  the number of land management professionals which are capable of distinguishing taxa at these resolutions, and have time assign C-values, are limited (@kramer2015report, @ahrends2011conservation, @morrison2016observer). Other possible limiting factors are that the FQI indices have been traditionally associated with the portions of Natural Resources focused on designation of parcels for conservation and preservation, rather than management of these parcels. 

While C-values exist for virtually all states East of the Continental Divide, Colorado is one of only two states with significant surface lands administered by the BLM which has existing C-values for every documented member of it's flora (@spyreas2019floristic). In fact BLM Colorado staff, including the lead state botanist, assisted in developing the states C-values (@cnhp2020fqi). Here we utilize FQA, to supplement our formal AIM analysis, and to develop a map of the habitat quality for lands managed by the Uncompahgre Field Office. 

The Floristic Quality (FQ) Assessments are comprised of two core indices Mean Coefficient of Conservatism (Mean CoC) and Floristic Quality Index (FQI) (see Appendix A for equations). While many novel permutations of these calculations exist, they seem to offer little additional insight to the main pair of indices and appear only useful for niche applications (@spyreas2019floristic). As the general goal of an FQA is the assessment of parcels of land, the location of study areas across different habitat types is an accepted use, and as we used a weighted stratified sample design our points meet the assumptions implicit in the sample design (@spyreas2019floristic). FQ Assessments have yet to converge on a standardized size for conducting the species inventory, and while Mean-CoC is affected by plot size FQI is relatively robust against small plot size. Regardless in similar systems plots of similar size as AIM plots have been shown to be adequate for noting enough species to calculate both main indices (@spyreas2016scale). In several applications C-values have been shown to be stable across sampling time, in part perhaps due to a propensity for many species at a site to share the same C-values (@spyreas2016scale, @matthews2015null, @bried2013floristic). C-values have also been shown capable of distinguishing habitat variability more effectively than two traditional diversity metrics (Simpsons and Shannons) (@taft2006estimating). Practitioners of varying degrees of skill are likely to have minimal effects on the estimates of the FQA indices due to the species encoding some degrees of redundant information (@bried2018experts, @spyreas2019floristic). Further, the calculations of Mean C and FQI, are relatively independent of the species richness, or diversity at any one plot, and are capable of reflecting the C-values of the species present, rathern than other diversity metrics *per se*.

#### Utility and Limitations of FQA
FQA scores are tied to the regional list of C-values, accordingly they cannot be compared across regions with independently developed C-values, in other words the FQA scores of sites in the mixed grass prairies of Kansas and Colorado are incomparable. Scores have the potential to be misleading if compared across major habitats, (e.g. comparing Sagebrush to Salt desert) but generally appear robust against this (@spyreas2019floristic). The indices are relatively boundless, e.g. we could visit plots which we designate high quality and use them as benchmark for FQI values, but we cannot incorporate metrics e.g. land > 4 is 'good', until we perform these activities to determine locally relevant measurements. 

> _'... tolerance of anthropogenic disturbance and exclusivity to remnant habitats are the only validated criteria for deﬁning FQA.'_
> 

> _"...FQA conveys two things about high conservative species: (1) All else being equal, they have greater conservation value, and (2) they reﬂect a site’s history of minimal disturbance and degradation."_
>
> `r tufte::quote_footer('--- Spyreas 2019')`

Accordingly, in our scenario we are interested in using FQA to identify sites which have histories of evident degradation. 

## Methods

\begin{wrapfigure}{r}{0.65\textwidth}
  \centering
    \includegraphics[width=0.65\textwidth]{../plots/maps/FQI-plots.png}
  \caption{FQI Values observed at AIM Plots}
\end{wrapfigure}

Cleaned AIM species richness data were imported from TerrAdat and joined to the CNHP C-values using the lookup keys developed for the Functional Diversity in Section 11. All species from the plot based species richness which were not unambiguously identified to terminal taxon were removed from analysis. The Mean Coefficient of Conservatism and Floristic Quality Index were calculated via the formulation in Appendix 1. To determine whether their were bias in the Mean-CoC values between the strata, a linear model with a single set categorical predictors and a single continuous response (an 'ANOVA', or Analysis of Variance) was used, with a Kruskall-Wallis Post-hoc test with 95% Confidence Intervals. The strata used for detecting differences were those developed in **SECTION XXX**.

```{r load libraries and define functions}
library(sf)
library(tidyverse)
library(ggpubr)
source('functions.R')
mcoc <- function(x){sum(x)/length(x)} 
fqi <- function(x){sum(x)/length(x) * sqrt(length(x))}

rm(convex_hull_resample)
```

```{r Import AIM Species Richness Data}
praw <- '../data/raw'
ppro <- '../data/processed'
f <- list.files(praw, pattern = 'csv')
fp <- list.files(ppro, pattern = 'csv')

spp_richness <- read.csv(file.path(praw, f[grep('SpeciesRichness[.]', f)])) %>% 
  dplyr::select(SpeciesList, PrimaryKey)
spp_r_header <- read.csv(file.path(praw, f[grep('RichnessHeader', f)])) %>% 
  dplyr::select(FormDate, Observer, PrimaryKey)

pts <- st_read(
  '../../aimDB/data/raw/AIM_Sample_Design/AIM_Design_Stratification.shp',
                quiet = T) %>% 
  st_transform(26913) %>% 
  st_buffer(65) %>% 
  dplyr::select(PLOTID, STRATUM) 

char <- read.csv(file.path(praw, f[grep('Characterization', f)] ) ) %>% 
  st_as_sf( coords = c('Longitude', 'Latitude'), crs = 4269) %>% 
  filter(str_detect(PrimaryKey, negate = T,
                    'Fuels|Bull-Draw|CO-NWD|_Tres_Rios|Grand_Junction|Gunnison|Moab|TRFO'),
         str_detect(PlotID, '[A-Z]{2,3}-[0-9]{2,3}')) %>% 
  st_transform(26913) %>% 
  dplyr::select(PrimaryKey) 

spp_richness <- st_intersection(pts, char)  %>% 
  left_join(., spp_r_header, by = 'PrimaryKey', multiple = "all") %>% 
  left_join(., spp_richness, by = 'PrimaryKey', multiple = "all") %>% 
  rename(SYMBOL_AIM = SpeciesList)  %>% 
  dplyr::select(PLOTID, STRATUM, SYMBOL_AIM, geometry)

no_records_full <- nrow(spp_richness)

rm(pts, char, spp_r_header)
```
 
```{r Join C Values to Species Richness}

# remove unknowns here, only 118 records total across 5 years, pretty impressive!
unks <- spp_richness %>% 
  filter(str_detect(SYMBOL_AIM, '^AG|^PG|^AF|^PF|^SH\\d+')) %>%
  nrow()

spp_richness <- spp_richness %>% 
  filter(str_detect(SYMBOL_AIM, '^AG|^PG|^AF|^PF|^SH\\d+', negate = T))

# add on the official USDA codes to the AIM codes to sync up with CNHP
attributes <- read.csv(
  file.path(ppro, fp[grep('Attribute.*Table-RCB', fp)]) ) %>% 
  dplyr::select(SYMBOL_AIM, SYMBOL_USDA) %>% 
  distinct(.keep_all = T)   # one duplicate in here somewhere!

spp_clean <- left_join(spp_richness, attributes, by = 'SYMBOL_AIM', multiple = "all") %>% 
  mutate(SYMBOL_USDA = if_else( is.na(SYMBOL_USDA), SYMBOL_AIM, SYMBOL_USDA))

cvals <- read.csv(file.path(ppro, fp[grep('C-Values', fp)])) %>% 
  dplyr::select(SYMBOL_USDA = National_USDASymbol, C.Val = FQA_C.Value2020_Numeric, Native = FQA_NativeStatus, 
         Ack_SciName_noAuthority)

spp_clean <- left_join(spp_clean, cvals, by = 'SYMBOL_USDA', multiple = "all") %>% 
  # due to multiple look ups we will remove duplicate records
  group_by(PLOTID, SYMBOL_USDA) %>% 
  slice_sample(n = 1)

# spp_clean %>%  # check for duplicates here. 
#  group_by(PLOTID, SYMBOL_USDA) %>% 
#  filter(n() > 1)

no_CVAL_match <- filter(spp_clean, is.na(C.Val)) %>% 
  nrow()

spp_clean <- drop_na(spp_clean) %>% 
  group_by(PLOTID)

rm(cvals, attributes)
```

```{r Calculate Coefficients}
FQI_results <- spp_clean %>% 
  summarize(mcoc_r = mcoc(C.Val), 
            fqi_r = fqi(C.Val)) %>% 
  st_as_sf()

rm(fqi, mcoc)
```

```{r export coefficients for mapping, eval = F}

st_write(FQI_results, '../results/FQI_values.shp')
```

```{r add strata to site}

lc_lkp <- read.csv(file.path(praw, f[grep('landcover', f)]) )  %>% 
  cbind(., RealStratum = c('MC', 'AS', 'MMS', 'PJ', 'SD', 'SS'))
r <- terra::rast('../../plot_post_stratification/data/processed/PredictedVegClass.tif')
pts <- FQI_results %>% 
  dplyr::select(PLOTID) %>% 
  terra::vect() %>% 
  terra::project(terra::crs(r))

reclass <- terra::extract(r, pts, method = 'simple', bind = T) %>% 
 as.data.frame() %>% 
  mutate(lyr1 = case_when(
         PLOTID %in% c('SS-313-', 'PJ-143', 'PJ-153', 'PJ-164', 'SS-313') ~ 4, 
         PLOTID %in% c('AS-002', 'AS-004', 'MMS-097') ~ 3, 
         PLOTID %in% c('MC-068', 'AS-012') ~ 2, 
         PLOTID %in% c('SS-306') ~ 6,
         TRUE ~ lyr1)) 
         
FQI_results <- left_join(FQI_results, reclass, by = 'PLOTID') %>% 
  left_join(., lc_lkp, by = c('lyr1' = 'layer')) %>% 
  dplyr::select(-lyr1) %>% 
  relocate(stratum, .before = geometry) %>% 
  st_as_sf() %>% 
  mutate(stratum = str_trim(stratum)) 

rm(r, pts, reclass, lc_lkp)
```

```{r determine whether the metrics vary by classified stratum}

wghts <- read.csv( file.path(praw, f[grep('Weights', f)])) %>% 
  dplyr::select(WghtPerPlot, STRATUM = Stratum, AoInference = Area)

fqi_weights <- read.csv(file.path(praw, f[grep('Interence', f)])) %>% 
  dplyr::select(PLOTID = Plot.ID, AoInference, xcoord, ycoord) %>% 
  right_join(., FQI_results, by = 'PLOTID') %>% 
  rowwise() %>% 
  mutate(STRATUM = str_remove(str_extract(PLOTID, '^.*-'), '-')) %>% 
  left_join(., wghts, by = c('STRATUM','AoInference'))

rm(wghts)
```

```{r Plot difference of MCOC by stratum 1}

fqi_weights_sub <- fqi_weights %>% filter(RealStratum != 'AS')

fqa_meds <- fqi_weights_sub %>% 
  group_by(RealStratum) %>% 
  summarize(med_C = median(mcoc_r),
            mean_C = mean(mcoc_r))
```

```{r Plot difference of MCOC by stratum 2, eval = F}
boxplot_drawer(df = fqi_weights_sub, response = mcoc_r, 
                     group = RealStratum, col_pal = strata_pal) +
  labs(title = 'Mean CoC by Stratum', x = 'Stratum', y = 'Mean Coefficient of Conservatism') 

ggsave(plot = last_plot(), device = 'png', height = 9, units = 'cm', 
       path = '../plots/graphics', filename = 'Mcoc_boxplot.png')
```

```{r}
rm(fqi_weights_sub, theme_boxplot, boxplot_drawer, strata_pal)
```

The floristic quality assessment data are theoretically independent of ecological sites. Accordingly, we can create a statistical model using the plots which were sampled, and using this model predict the values of floristic quality across the field office onto a map. Based on Figure 1, and our field experience collecting these data, we suspected that the variables which most strongly predict Mean-CoC scores, and which we could readily acquire or create were: 1) distance to nearest road, 2) patch size of federal public lands, 3) human population density within ~ 10 km (~6 miles), 4) elevation. These analyses occurred across the extent of the mapped area in Figure 1), at a resolution of 90 meters. All of these variables were used as predictors of Mean-CoC scores in our statistical model. 

Predictor variables were created using the following processes. To calculate the distance of each pixel of BLM land from the nearest road, the U.S. Census Bureau's 'roads' data set was acquired through 'tigris' and simplified using 'st_simplify'; the distance function of 'terra' was used to calculate the distance from each pixel to the nearest road (@walker2022tigris, @hijman2022terra, @pebesma2018sf). The U.S. Census Bureau roads data set contains nearly all of 'major' dirt and gravel roads across the field office, included many used for historic Uranium mining activities. To calculate the patch size of federal lands, i.e. all area across BLM, Forest Service, and National Park Service management areas etc., federal lands were queried from the PAD-US database and then areas dissected by roads were erased; the area of each parcel were then calculated using 'st_area' from sf, and converted from meters squared to hectares (@gergely2013pad). The patches were then converted into a raster using 'rasterize' from terra, and the size was saved as an attribute. To create estimates of population density within a distance of ~ 10km of each pixel of BLM Land, population density data at 30m resolution were downloaded from HDX, and summed to 90m resolution; these values were then summed using rolling windows in terra (@tiecke2017mapping). 90 meter resolution elevation data were acquired from EarthEnv (@tuanmu2014global). All data sets were cropped and re-sampled to a template to ensure optimal alignment of raster cells. 

```{r Predict FQI across field office - prepare covariates, eval = F}
library(terra)
library(tigris)

citation('tigris')

rp <- '../../AIM_Field_rasters'
rf <- list.files(rp, recursive = T, pattern = 'tif')
p2carto <- '../../UFO_cartography'

vector_data <- list.files(p2carto, recursive = T, pattern = 'shp$')
popDen28 <- rast(file.path(rp, rf[grep('population.*28', rf)] ))
popDen38 <- rast(file.path(rp, rf[grep('population.*38', rf)] ))

# import extent file
ext <- vect('../../UFO_cartography/UFO_mask/mask/mask.shp') 

ext_84 <- ext %>% 
  project(crs(popDen28)) %>% 
  ext()

## import elevation data at 90m resolution
elev90 <- rast(file.path(p2carto, list.files(p2carto, recursive = T, pattern = 'DEM.*tif')))
elev90 <- crop(elev90, ext)

################          Human population Density          ####################
# data were downloaded on Feb 21, 14:15 CST, from (HDX)[https://data.humdata.org/dataset/united-states-high-resolution-population-density-maps-demographic-estimates] 
# for the area 38N - 48N, 100 - 110W. These data were cropped and masked to the outer extent of the UFO field office. 

# crop to extent of analysis
popDen <- mosaic(popDen28, popDen38, fun = '
                 ', overwrite = T,
                 filename = file.path(rp, 'PopDensity_mosaic.tif'))

popDen <- crop(popDen, ext_84, mask = T, overwrite = T)

# aggregate to 90m resolution
popDen_c <- aggregate(popDen, fact=3, cores=20, fun = 'sum', na.rm = T)

# create a rolling window SUM, over 9990m (~10 km area)
popDen_c <- focal(popDen_c, w = 111, fun = 'sum', expand = T, overwrite = T, na.rm = T)

# project into planar coordinates
popDen_c <- project(popDen_c, crs(ext), overwrite = T, filename = 
                      file.path(rp, 'PopDensity_planar.tif'))
popDen_c <- resample(popDen_c, elev90, overwrite = T, filename = 
                       file.path(rp, 'PopDensity_planar_resamp.tif'))

rm(popDen, popDen28, popDen38, ext_84)

####################### distance from nearest road #############################
# ensure that roads data are in planar coordinates systems before converting
# simplify roads data 

ext_sf <- st_read('../../UFO_cartography/UFO_mask/mask/mask.shp')

CO <- counties(state = 'CO', cb = T)  %>%
  dplyr::select(COUNTYFP, NAME ) %>% 
  st_transform(st_crs(ext_sf))
counties <- st_join(ext_sf, CO) %>% pull(COUNTYFP)

CO_roads <- roads(state = 'CO', county = counties, keep_zipped_shapefile = T) %>% 
  st_transform(st_crs(ext_sf)) %>% 
  st_crop(., ext_sf) %>% 
  st_simplify() %>% 
  dplyr::select(geometry) %>% 
  st_cast('LINESTRING') %>% 
  st_as_sfc()

rm(CO, counties)

##  now create distance to nearest roads  raster  ###

CO_roads_v <- vect(CO_roads)
road_raster <- rasterize(CO_roads_v, elev_90,
            background = NA)
m90_r <- distance(road_raster)
writeRaster(m90_r, file.path(rp, '90m_RoadDistance.tif'), overwrite = T)

################# patch size of contiguous public land - PADUS #################
padus <- st_read(
  file.path(p2carto, vector_data[grep('PAD.*Fee*', vector_data)]), quiet = T) %>% 
  filter(Mang_Type == 'FED') %>% 
  group_by(Mang_Type) %>% 
  summarize(geometry = st_union(geometry))#

# now erase the roads from it 

CO_roads_buff <- CO_roads %>% 
  st_as_sf() %>% 
  st_buffer(., 20) %>% 
  st_make_valid() %>% 
  st_union() %>% 
  st_as_sf() %>% 
  st_simplify()

padus_erased <- rmapshaper::ms_erase(padus, CO_roads_buff)

padus_erased <- padus_erased %>% 
  st_cast(., to = "POLYGON")  %>% 
  mutate(AreaHA = as.numeric(st_area(.)) / 10000 ) 

patchsize_r <- padus_erased %>% 
  vect() %>% 
  rasterize(., elev90, 'AreaHA',
            background = 0)

writeRaster(patchsize_r, file.path(rp, 'public_patchsize.tif'), overwrite = T)

rm(time90m, flat_lines_psp, L, CO_roads, m90_r, CO_roads_v, road_raster,
   padus_erased, CO_roads_buff, patchsize_r)
```

```{r prepare raster stack for FQI prediction}
library(terra)

rp <- '../../AIM_Field_rasters'
rf <- list.files(rp, recursive = T, pattern = 'tif')
p2carto <- '../../UFO_cartography'

ext <- vect('../../UFO_cartography/UFO_mask/mask/mask.shp') 

elev_90 <- rast(file.path(p2carto, list.files(p2carto, recursive = T, pattern = 'DEM.*tif')))
elev_90 <- crop(elev_90, ext)

popDen <- rast(file.path(rp, rf[grep( 'PopD.*resamp', rf)]))
roadDist <- rast(file.path(rp, rf[grep( 'RoadDistance', rf)]))
patch <- rast(file.path(rp, rf[grep( 'public_patchsize.tif', rf)]))

popDen <- crop(popDen, elev_90)
popDen <- resample(popDen, elev_90)
popDen[is.na(popDen[])] <- 0
patch <- crop(patch, elev_90)
patch <- resample(patch, elev_90)
roadDist <- crop(roadDist, elev_90)
roadDist <- resample(roadDist, elev_90)

# make X and Y coords explicit layers in the model. 
x_rast <- rast(elev_90)
x_rast <- init(x_rast, "x")

y_rast <- rast(elev_90)
y_rast <- init(y_rast, "y")

pred_stack <- c(elev_90, popDen, patch, roadDist ) #,
             #   x_rast, y_rast)
names(pred_stack) <- c('elevation', 'pop_density', 'patch_size', 'road_dist') #,
                    #   'xcoord', 'ycoord')

rm(ext)
```

```{r Dredge a consensus model}

# Run automated model selection 
library(MuMIn)
library(usdm)

# check for correlation among the independent variables from 5000 random cells
corr_check_sample <- terra::spatSample(pred_stack, size = 5000, method = 'regular')
vif_results <-  usdm::vifstep(corr_check_sample, th=10)
# vif_results

rm(corr_check_sample, vif_results)

## extract values to AIM Plots
fqa_results_V <- FQI_results %>% 
  dplyr::select(PLOTID, mcoc_r) %>% 
  vect()
fqa_rast_vals <- extract(pred_stack, fqa_results_V,  bind = T) %>%
  data.frame()

# pull off plotID
fqa_rast_vals <- dplyr::select(fqa_rast_vals, -PLOTID)
fqa_rast_vals <- fqa_rast_vals[complete.cases(fqa_rast_vals),]

## attempt to model FQA as a function of human disturbance, and elevation. 

# this is the theoretical full model:
# glm(mcoc_r ~ road_dist * pop_density * patch_size * elevation * xcoord * ycoord, data = fqa_rast_vals)

oop <- options(na.action = "na.fail")

full_model <- glm(mcoc_r ~ ., data = fqa_rast_vals)
optimal_models <- dredge(full_model, evaluate = T, rank = AICc)
# optimal_models

# gather the actual terms from the top performing models, to see what worked.
model_terms <- get.models(optimal_models, subset = T)

# if top models have dAIC < 2.0, than 'stack' the models
stacked_model <- model.avg(optimal_models, subset = delta < 2.1, fit = T)

# these provides the estimates of our coefficients
# stacked_model$coefficients
# identify out top component models
# stacked_model

rm(full_model)
```


To determine whether any predictor variables would violate assumptions of independence, by being collinear, variance inflation factors were calculated using the 'function' vifstep in the package usdm with a theta cut off 10 (@naimi2014usdm). This function was used on a subset of 5000 random pixels regularly dispersed across the extent of the sample area, which had their value of all predictors extracted from them. No theta scores were high enough that they suggested any combination of variables should be removed from the analysis, but notable collinearity existed between 'elevation' and 'X coordinates'; this is likely due to the increase in elevation from the Colorado Plateau into the Rocky Mountains, which follows a longitudinal (West to East) trend. 

In order to create a model which could capture the variation in our data and generalize them to predict Mean-CoC, a maximal model of the term _glm(mcoc_r ~ road_dist * pop_density * patch_size * elevation * xcoord * ycoord)_ was created. This model was then fed into the 'dredge' function from the package MuMIn, which creates all smaller models, and evaluates them in an information theoretic framework (@Barton2022mumin). All tops models, those with $\Delta$AIC < 2.0, were each selected for further analyses. Each model was checked for the effect of spatial autocorrelation in the residuals by identifying neighboring points using 'graph2nb'  and converting them into three neighbor lists using 'nb2listw' (both functions from 'spdep'), the Moran's Index for each model was very low, despite the Moran's Index for the predictor variables being very high and showing significant clustering (@bivand2022spdep). Subsequent to the interpretation of the maps produced by the stacked model outputs, considering that the model including the coordinates did not suffer from autocorrelation, and that the X coordinate was often selected as a predictor in top models, we believe that minor collinearity between the X coordinate and the strongest predictor Elevation indicated that the term was redundant. 

A new full model, without the coordinates, was created in the form _glm(mcoc_r ~ road_dist * pop_density * patch_size * elevation)_ and passed to the 'dredge' function in MuMIn. The results of the top models, those with $\Delta$AIC <= 2.13 were checked for spatial autocorrelation, using the same methods as the models above. The general rule of thumb for only utilizing models with $\Delta$AIC < 2, was bent in order to accommodate two more models which incorporated terms we have observed to be ecologically relevant, and are well supported in the landscape ecology primary literature (@symonds2011brief). After evaluation of the stacked prediction map from these top models, it was used for the final analysis, in part because removal of the x-coordinates made it more easy to interpret, and it was able to utilize the other predictors which are known to affect species compositions, both universally and strongly. 

All six of the optimal models, which were in the ensemble model, were individually predicted into space and the weighted means, based on each models AIC weight, were used to generate an ensembled prediction layer (@symonds2011brief, @Barton2022mumin, @dormann2018model). Spatial predictions of the statistical model were performed using terra @hijman2022terra. The ensembled prediction was then clipped to the extent of UFO BLM administered lands. 

```{r report top models, eval = F}
summary(stacked_model)
sw(optimal_models)
```

```{r stack the top models}

# we will create predictions from EACH model with dAIC < 2.0 and then take the m
# mean of each model 

# model_terms[["6"]][["call"]]
# model_terms[["8"]][["call"]]
# model_terms[["2"]][["call"]]
# model_terms[["4"]][["call"]]
# model_terms[["14"]][["call"]]
# model_terms[["16"]][["call"]]

model_6 <- glm(formula = mcoc_r ~ elevation + pop_density + 1, data = fqa_rast_vals)
model_8 <- glm(formula = mcoc_r ~ elevation + patch_size + pop_density + 1, data = fqa_rast_vals)
model_2 <- glm(formula = mcoc_r ~ elevation + 1, data = fqa_rast_vals)
model_4 <- glm(formula = mcoc_r ~ elevation + patch_size + 1, data = fqa_rast_vals)
model_14 <- glm(formula = mcoc_r ~ elevation + pop_density + road_dist +  1, data = fqa_rast_vals)
model_16 <- glm(formula = mcoc_r ~ elevation + patch_size + pop_density + road_dist + 1, data = fqa_rast_vals)

top_pred_r <- c(
  pred_6 <- predict(pred_stack, model_6, type="response"),
  pred_8 <- predict(pred_stack, model_8, type="response"),
  pred_2 <- predict(pred_stack, model_2, type="response"), 
  pred_4 <- predict(pred_stack, model_4, type="response"), 
  pred_16 <- predict(pred_stack, model_14, type="response"), 
  pred_18 <- predict(pred_stack, model_16, type="response")
)

# plot(top_pred_r) 
# create a simple ensemble, by calculating the mean of each top model
top_pred_r <- weighted.mean(top_pred_r, w = stacked_model$msTable$weight)
# plot(top_pred_r)

rm(model_6, model_8, model_2, model_4, model_16, model_14, oop, pred_16, 
   pred_2, pred_4, pred_6, pred_8, pred_18, model_terms)
```

```{r test for spatial autocorrelation, eval = F}

fqa_rast_vals <- extract(pred_stack, fqa_results_V,  bind = T) %>%
  data.frame() %>% 
  dplyr::select(-PLOTID)

pred_stack <- c(elev_90, popDen, patch, roadDist, x_rast, y_rast)
names(pred_stack) <- c('elevation', 'pop_density', 'patch_size', 'road_dist',  'xcoord', 'ycoord')

xy <- data.matrix(fqa_rast_vals[,c("xcoord", "ycoord")])
nb.gab <- spdep::graph2nb(spdep::gabrielneigh(xy), sym=TRUE)
par(mar=c(0,0,0,0))
plot(nb.gab, xy)
listw.gab <- spdep::nb2listw(nb.gab)

dlist <- spdep::nbdists(nb.gab, xy)
dlist <- lapply(dlist, function(x) 1/x)
listw.d1 <- spdep::nb2listw(nb.gab, style = "W", glist=dlist)
dlist <- lapply(dlist, function(x) 1/x^2)
listw.d2 <- spdep::nb2listw(nb.gab, style = "W", glist=dlist)

apply(fqa_rast_vals, MAR = 2, spdep::moran.test, listw.gab)
apply(fqa_rast_vals, MAR = 2, spdep::moran.test, listw.d1)
apply(fqa_rast_vals, MAR = 2, spdep::moran.test, listw.d2)

m_list <- list(model_6, model_8, model_2, model_4, model_16, model_14,  model_16)
lapply(m_list, spdep::lm.morantest, listw.gab) 
lapply(m_list, spdep::lm.morantest, listw.d1) 
lapply(m_list, spdep::lm.morantest, listw.d2) 

rm(xy, nb.gab, listw.gabv, listw.d1, listw.d2, m_list, listw.gab, dlist)
```

```{r remove some modelling variables}
rm(fqa_rast_vals, pred_stack, x_rast, y_rast, fqi_weights, 
   elev_90, popDen, patch, roadDist, fqa_results_V)
```

```{r Crop prediction of FQI to UFO land, eval = F}

p2carto <- '../../UFO_cartography'
vector_data <- list.files(p2carto, recursive = T, pattern = 'shp$')
ufo_land <- vect(file.path(p2carto, vector_data[grep('*Surface*', vector_data)])) |>
  aggregate()

top_pred_crop <- crop(top_pred_r, ufo_land, mask = T, overwrite = T, 
                      filename = '../results/predicted_FQI_noXY.tif')

top_pred_crop <- crop(top_pred_r, ufo_land, mask = T, overwrite= T,
                      filename = '../results/predicted_FQI.tif')

rm(ufo_land, p2carto, vector_datatop_pred_r, top_pred_crop)
```

```{r}
rm(top_pred_r)
```

# Discussion

```{r helper function}

my_range <- function(x, col){
  paste(round(min(x[,col]),2), '-', round(max(x[,col]),2) )
}
```

\begin{wrapfigure}{r}{0.5\textwidth}
  \centering
    \includegraphics[width=0.5\textwidth]{../plots/graphics/Mcoc_boxplot.png}
  \caption{Comparision of median values by Stratum}
\end{wrapfigure}

The Mean-CoC values varied by stratum (Analysis of Variance (ANOVA)), follow-up tests (Kruskall-Wallis) indicated that Salt Desert and Sagebrush sites did not differ from each other, nor Mixed Mountain Shrub and Pinon-Juniper from each other (Figure 1). Results indicate that all members of these two groups differed across them, e.g. Sagebrush differed from both Mixed Mountain Shrub and Pinon-Juniper and *vice versa*.  The two lower elevation strata, Salt Desert and Sagebrush, are generally more accessible and have higher recreational land uses, and had the lowest Mean-CoC. While the values did vary, the extent of the variation was minor, with the range of variation from the median values `r my_range(fqa_meds, 'med_C')` and a mean range of `r my_range(fqa_meds, 'mean_C')`, given that these values occur over a range of 0-10, we considered this to be a negligible difference. Which was not indicative of major biases in the C-values ascribed to plants themselves, but rather indicative of actual land conditions, and that all main vegetation types in the UFO could be analysed together.

\begin{wrapfigure}{r}{0.6\textwidth}
  \centering
    \includegraphics[width=0.6\textwidth]{../plots/maps/FQI-prediction.png}
  \caption{Consensus predictions of FQI Values across the UFO Field Office (excluding lands East of Paonia along the North Fork of the Gunnison}
\end{wrapfigure}

The generalized linear models were used to predict the floristic quality of un-sampled areas on BLM Land. Based on AIC model selection a total of `r nrow(stacked_model$msTable)` top models were retained, and these were then combined into a single model, of which the weights used to make a stacked ensembled prediction of the Mean-CoC. The results from these top models indicate that all predictors, elevation, human population density, Federal land patch size, and distance to nearest road, affect the Mean-CoC, but that by far the strongest predictor is elevation (Figure 4, Appendix). Elevation was such a useful predictor, that under certain statistical paradigms (e.g. forward or backward selection via null hypothesis) our models may be simplified to include it as the only predictor. However, given that C-values reflect the tolerance of a species to anthropogenic disturbances this approach would be questionable. Rather we suspect that elevation is strongly associated with disparities in land use intensity, but more importantly also the inability of lower elevation sites to recover from previous disturbances relative to higher elevation areas. The other predictors had the expected effect on Mean-CoC as expected, Mean-CoC slowly increases as nearby human population density decreases, and FQI slowly increases as the size of the patch of natural lands and the distance from the nearest road increase. 

From the predicted Mean-CoC values we can infer that habitat quality follows a consistent trend across the field office (Figure 2). In general, the lowest elevation sites display the lowest values of FQI and Mean-CoC. This is especially apparent between Montrose and Grand Junction on Highway 50. These and several other areas of the Salt Desert along Highway 50, and near Crawford are areas of concern. However, not all Salt Desert is inherently in this condition as can be seen from the values for a moderately sized parcel of BLM managed land immediately Southeast of Montrose; indicating that both anthropogenic disturbances and elevation in these areas are causal factors for these low values. Areas with low FQI values represent highlight potential restoration needs and the Mean-CoC prediction model can be a tool to help prioritize such actions. The highest elevation portions of BLM managed land are in the best habitat condition, and very good conditions range into Pinon-Juniper, including most of the disturbed areas around historic mining in the West End. By these metrics considerable portions of the sagebrush habitat occupied by Gunnison sage-grouse near Crawford are also in good condition. 

Subjective interpretation suggests these results seem very similar to those generated by the much more time intensive comparison of Ecological Sites to Quantitative Benchmarks. This relationships considers serious consideration in the use of FQA as a proxy of site condition, and may warrant applications for rapid surveys prior to certain land use decisions, in time frames when AIM cannot be implemented. It also provides a second independent set of data which may help to contextualize sites which were unable to have their Ecological Sites correlated, are lacking Ecological Site Descriptions, or are difficult to interpret.

```{r remove the remaining variables}
rm(f, fp, no_CVAL_match, no_records_full, p2carto, ppro, praw, rf, rp, unks, my_range)
```

\newpage

## Appendix A - Indices

<br>    
<br>     

**Mean Coefficient of Conservatism** 
<br>       

$$
\overline{C} = \frac{\sum{} C_i}{S}
$$

Where:  
\begin{enumerate}[nosep]
  \item[] $\overline{C}$ is the Mean Coefficient of Conservatism, or for short Mean C
  \item[] $S$ is the number of species included in the calculation  
  \item[] $C_i$ in particular $C$ is the Conservatism Value (C-Value), for each $_i$ of the $S$ at the site  
  \item[] $\sum{}$ is an operator, meaning that we will calculate the sum of all C-Values, $C$  
\end{enumerate}

<br>    
<br>      

**Floristic Quality Index** 
<br>    

$$
FQI = \overline{C} * \sqrt{S}
$$

Where:  
\begin{enumerate}[nosep]
  \item[] $\overline{C}$ is the Mean Coefficient of Conservatism, or for short Mean C  
  \item[] $\sqrt{S}$ is the square root of the number of species included in the calculation  
\end{enumerate}

<br>    
<br>    
<br>    
Equations from @swink1994plants, and modified for simpler formulations. 

```{r optimal_models, out.width="50%", fig.cap= 'Contributions of each term to each model, and each model sized by contributions to ensemble prediction. Note the left-most column is not a predictor, and only the top 6 models were used.', fig.align="center"}

plot(optimal_models, col.mode = 1, 
     vlabels = parse(text = paste("omega ==", round(Weights(optimal_models), 2))))
rm(stacked_model)
```


\newpage

## References

\small
 
