---
title: "Rare Species"
author: NULL
date: NULL
output:
  pdf_document: default
  word_document: default
header-includes:
 - \usepackage[width=\textwidth]{caption}
always_allow_html: yes
csl: ../citations/citations/apa.csl
bibliography: ../citations/citations/citations.bib
link-citations: yes
---

While the overall focus of the Assess, Inventory and Monitor Program is at a broad scale and seeks to be indicative of general trends of landscape status, the Inventory work has resulted in the fine resolution mapping and recording of many finer scale processes. In particular a number of rare plant occurrences have been noted over the life of the project in the UFO. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(dpi = 300) 
knitr::opts_chunk$set(message = F)
knitr::opts_chunk$set(warning = F)
```



```{r Import AIM species richness}
praw <- '../data/raw'
f <- list.files(praw, pattern = 'csv')

richness <- read.csv(file.path(praw, f[grep('Associations', f)])) %>% 
  filter(PHASE == '1.1', str_length(SYMBOL) >= 4)  
```


```{r Import tracked Plants}

read.csv(file.path(praw, f[grep('CNHP', f)]))  %>% 
  filter(MAJORGROUP == 'Vascular Plants')

```

