---
title: "Plant Species Diversity"
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
header-includes:
- \usepackage[width=\textwidth]{caption}
- \usepackage{wrapfig}
always_allow_html: yes
csl: ../citations/citations/apa.csl
bibliography: ../citations/citations/citations.bib
link-citations: yes
---

\vspace{-1cm}

In this section we discuss two very simple forms of diversity. *Alpha-diversity* - or species richness, is simply the number of species in a space at a point in time. The scale at which we discuss $\alpha-diversity$ is usually up to a few footballs in size. This metric refers to areas which we are intimately familiar with and may traverse readily on foot. *Gamma-Diversity*, represents the richness of species in a larger area, generally a landscape. For example, we may readily discuss the $\gamma-diversity$ of the Dominguez-Escalente National Monument. In all instances the $\alpha-diversity$ of sites are nested within the gamma diversity of an area. $\gamma-diversities$ -in this case the number of species- exceed those of $\alpha  -diversities$, both due to the relative uncommonness of many species, these uncommon species are often not present across the entirety of the landscape, and due to the large changes in the type of species supported by differing habitat types. These turnovers across the alpha diversity of sites, the difference in species present at sites, comprise *beta-diversity*. High rates of $\beta-diversity$, or dissimilarity of sites, foster high rates of gamma-diversity (@whittaker1972evolution).

Evolution, the process largely mediating the maintenance of diversity, is survival of the fittest. However, the conditions of the test which may constitute the 'fittest' are nearly as numerous as the forms of diversity. Many species which exist in the same location in space, have distinct characteristics which allow for species persistence; dry years favor some species, while wet years favor others, some require more sun, while others thrive with less, and there are myriad permutations and combinations of these settings. The cover of these species ebb and flow with the usual weather and disturbances within the climate zone of the site (@hoover2014resistance, ...). These trends are especially important for the production of forage and browse, over the life of most large animals, they will have to feed on what they have available. No single species is the fittest at a site on a time scale which the BLM manages land, having multiple species which function similarly in space is the only stable strategy for management.

While Ecological Site Descriptions do not provide true measures of $\alpha-diversity$ many of them do contain lists of taxa which may be considered *dominant* or *subordinate* at a site (@avolio2019demystifying, @grime1998benefits). In general, it seems that the Shrubs, Trees, and Grasses at an Ecological Site would be considered *dominant*, and the forbs *subordinate*; keep in mind these terms refer to immediate ecosystem cycling effects, and the services offered by forbs to insects and then larger animals remain substantial (@avolio2019demystifying). These species which have high amounts of biomass, and ground cover, may be thought of as a core groups of species which are essential for the functioning of an ecosystem (@grime1998benefits), and each of the dominant species have been theorized to have conceptual effects as large as their cover. When dominant and subordinate species are lost from an area, it also has effects on the remaining species - most of which are relatively uncommon across the landscape (@grime1998benefits, @whittaker1965dominance). Considerable research has shown that having a diverse suite of plant species allows areas to: 1) Produce more forage in both a single year, and across different weather scenarios (@vogel2012grassland, @hoover2014resistance), 2) recover from disturbances such as fire, or compaction (@tilman1994biodiversity) 3) and resist degradation such as from the encroachment of noxious weeds (@weisser2017biodiversity, @avolio2019demystifying, @allan2011more, @gaitan2014vegetation, @sheley2010resistance, @isbell2011high, @oakley2013plant, *and reviewed in* @maestre2016structure, @oliver2015biodiversity). 

Recruitment of native species back following drought may have issues pertaining to seed recruitment... @tilman1992drought .. although soil seeds banks in arid systems well.

Given the generally high % abundance of the taxa listed in the species tables ESD's it is unlikely that they are truly lost, and should at least show in species richness meanders which have higher probability of uncovering more microhabitats, areas which individuals may have more access to water or other ... @hooper2005effects


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(dpi = 300) 
knitr::opts_chunk$set(message = F)
knitr::opts_chunk$set(warning = F)
```

Information on production at Ecological Sites implies some superficial, yet essential, components of plant diversity. Here we determine what proportion of species identified in ESD production tables are present at each AIM plot, how many are missing, and whether any species are uniformly missing. We also combine plots by Ecological Site to determine $\gamma-diversity$ by site, and compare the relative turn over in species composition within each Ecological Site which has replicate plots.  

# Methods

```{r}
library(tidyverse)
library(sf)
library(cowplot)
```


```{r paths and files}
praw <- '../data/raw'
ppro <- '../data/processed'
f <- list.files(praw, pattern = 'csv')
fp <- list.files(ppro, pattern = 'csv')
source('functions.R')
rm(strata_pal, boxplot_drawer, convex_hull_resample, theme_boxplot)
```


```{r import aim data for species richness}

spp_richness <- read.csv(file.path(praw, f[grep('SpeciesRichness[.]', f)])) %>% 
  select(SpeciesList, PrimaryKey)
spp_r_header <- read.csv(file.path(praw, f[grep('RichnessHeader', f)])) %>% 
  select(FormDate, Observer, PrimaryKey)

pts <- st_read(
  '../../aimDB/data/raw/AIM_Sample_Design/AIM_Design_Stratification.shp',
                quiet = T) %>% 
  st_transform(26913) %>% 
  st_buffer(65) %>% 
  select(PLOTID, STRATUM) 

char <- read.csv(file.path(praw, f[grep('Characterization', f)] ) ) %>% 
  st_as_sf( coords = c('Longitude', 'Latitude'), crs = 4269) %>% 
  filter(str_detect(PrimaryKey, negate = T,
                    'Fuels|Bull-Draw|CO-NWD|_Tres_Rios|Grand_Junction|Gunnison|Moab|TRFO'),
         str_detect(PlotID, '[A-Z]{2,3}-[0-9]{2,3}')) %>% 
  st_transform(26913) %>% 
  select(PrimaryKey) 

spp_richness <- st_intersection(pts, char)  %>% 
  left_join(., spp_r_header, by = 'PrimaryKey', multiple = "all") %>% 
  left_join(., spp_richness, by = 'PrimaryKey', multiple = "all") %>% 
  rename(SYMBOL_AIM = SpeciesList) 

no_records_full <- nrow(spp_richness)

spp_richness <- spp_richness %>% 
  filter(str_detect(SYMBOL_AIM, '^AG|^PG|^AF|^PF|^SH\\d+', negate = T))

# add on the official USDA codes to the AIM codes to sync up with CNHP
attributes <- read.csv(
  file.path(ppro, fp[grep('Attribute.*Table-RCB', fp)]) ) %>% 
  distinct() # one duplicate in here somewhere!

Pk2PID <- spp_richness %>% 
  select(PrimaryKey, PLOTID) %>% 
  st_drop_geometry() %>% 
  distinct()

spp_richness <- left_join(spp_richness, attributes, by = 'SYMBOL_AIM') %>% 
  filter(INVASIVE == F) %>% 
  select(-FormDate, -Observer, -INVASIVE, -NATIVITY, -SYMBOL_AIM,
         SYMBOL = SYMBOL_USDA, -BINOMIAL_NAT, -PrimaryKey)

spp_richness_clean <- nrow(spp_richness)

rm(spp_r_header, char, pts)
```


```{r Import Data for LPI}

lpi <- read.csv( file.path(praw, 'LPIRAW.csv') ) %>% 
  select(PrimaryKey, RecKey, PointLoc:SoilSurface) %>% 
  mutate(across(.cols = TopCanopy:SoilSurface, ~ na_if(.x, ""))) %>% 
  pivot_longer(TopCanopy:SoilSurface, values_to = 'SYMBOL_AIM', names_to = 'Intercept') %>% 
  drop_na(SYMBOL_AIM) %>% 
  filter(SYMBOL_AIM != 'None', str_length(SYMBOL_AIM) >= 4) 

lpi <-  left_join(lpi, attributes, by = 'SYMBOL_AIM', multiple = "all") %>% 
  group_by(PrimaryKey, SYMBOL_USDA) %>% 
  add_count(name = 'hits') %>% 
  distinct(PrimaryKey, SYMBOL_USDA, .keep_all = T) %>% 
  drop_na(FUNCTIONAL) %>% 
  filter(INVASIVE == F) %>% 
  select(-RecKey, -PointLoc, -PointNbr, -Intercept, -SYMBOL_AIM, -BINOMIAL_NAT,
         -INVASIVE, -NATIVITY, SYMBOL = SYMBOL_USDA) 

lpi <- left_join(lpi, Pk2PID, by = 'PrimaryKey') %>% 
  drop_na(PLOTID) %>% 
  select(-PrimaryKey)

lpi <- lpi %>% 
  filter(str_detect(SYMBOL, '^AG|^PG|^AF|^PF|^SH\\d+', negate = T))

rm(Pk2PID)
```

```{r Create Finer Functional Groups}

spp_richness <- spp_richness %>% 
  relocate(any_of(c("PHOTOSYNTHETIC", "RESPROUT", "DURATION", 'GROWTHHABITSUB')), 
           .after = last_col()) %>% 
  unite(col = FUNCTIONAL_FINE, PHOTOSYNTHETIC:GROWTHHABITSUB, 
        sep = "-", na.rm = T, remove = F) %>% 
  select(-PHOTOSYNTHETIC, -RESPROUT) %>% 
  mutate(
    FUNCTIONAL_FINE = str_remove(FUNCTIONAL_FINE, 'CAM-'), 
    FUNCTIONAL_FINE = str_replace(FUNCTIONAL_FINE, 'SUCCULENT|SUBSHRUB', 'SHRUB'),
    FUNCTIONAL_FINE = str_replace(FUNCTIONAL_FINE, 'PERENNIAL-SHRUB', 'SHRUB'), 
    FUNCTIONAL_FINE = str_replace(FUNCTIONAL_FINE, 'PERENNIAL-TREE', 'TREE'), 
    FUNCTIONAL_FINE = str_replace(FUNCTIONAL_FINE, 'GRAMINOID|SEDGE', 'GRASS'),
    GROWTHHABITSUB = str_replace(GROWTHHABITSUB, 'SUCCULENT|SUBSHRUB', 'SHRUB'),
    GROWTHHABITSUB = str_replace(GROWTHHABITSUB, 'GRAMINOID|SEDGE', 'GRASS')) %>% 
  select(PLOTID, SYMBOL, BINOMIAL_ACK, FUNCTIONAL_FINE, GROWTHHABITSUB)

lpi <- lpi %>% 
   relocate(any_of(c("PHOTOSYNTHETIC", "RESPROUT", "DURATION", 'GROWTHHABITSUB')), 
           .after = last_col()) %>% 
  unite(col = FUNCTIONAL_FINE, PHOTOSYNTHETIC:GROWTHHABITSUB, 
        sep = "-", na.rm = T, remove = F) %>% 
  select(-PHOTOSYNTHETIC, -RESPROUT) %>% 
  mutate(
    FUNCTIONAL_FINE = str_remove(FUNCTIONAL_FINE, 'CAM-'), 
    FUNCTIONAL_FINE = str_replace(FUNCTIONAL_FINE, 'SUCCULENT|SUBSHRUB', 'SHRUB'),
    FUNCTIONAL_FINE = str_replace(FUNCTIONAL_FINE, 'PERENNIAL-SHRUB', 'SHRUB'), 
    FUNCTIONAL_FINE = str_replace(FUNCTIONAL_FINE, 'PERENNIAL-TREE', 'TREE'), 
    FUNCTIONAL_FINE = str_replace(FUNCTIONAL_FINE, 'GRAMINOID|SEDGE', 'GRASS'),
    GROWTHHABITSUB = str_replace(GROWTHHABITSUB, 'SUCCULENT|SUBSHRUB', 'SHRUB'),
    GROWTHHABITSUB = str_replace(GROWTHHABITSUB, 'GRAMINOID|SEDGE', 'GRASS')) %>% 
  ungroup() %>% 
  select(PLOTID, SYMBOL, BINOMIAL_ACK, FUNCTIONAL_FINE, GROWTHHABITSUB) 

```

```{r filter to native species}

nativity <- read.csv(file.path(ppro, fp[grep('Native', fp)])) %>% 
  select(SYMBOL, LIFECYCLE = DURATION, NATIVITY) %>% 
  mutate(across(.cols = everything(), ~ str_to_upper(.))) %>% 
  mutate(LIFECYCLE = case_when(
    LIFECYCLE == 'BIENNIAL, PERENNIAL' ~ 'PERENNIAL', 
    LIFECYCLE == 'ANNUAL, BIENNIAL' ~ 'ANNUAL', 
    LIFECYCLE == 'ANNUAL, PERENNIAL' ~ 'ANY',
    LIFECYCLE == 'ANNUAL, BIENNIAL, PERENNIAL' ~ 'ANY',
    TRUE ~ as.character(LIFECYCLE)
  )) %>% 
  distinct()

lpi <- left_join(lpi, nativity, by = 'SYMBOL', multiple = "all") %>%
  filter(NATIVITY == 'NATIVE') %>% 
  select(-NATIVITY)
  
spp_richness <- left_join(spp_richness, nativity, by = 'SYMBOL', multiple = "all") %>%
  filter(NATIVITY == 'NATIVE') %>% 
  select(-NATIVITY)

rm(nativity)
```


```{r gather estimate of number of species in functional groups}

dominant_plants <- read.csv(file.path(ppro, fp[grep('Associations', fp)])) %>% 
  filter(PHASE == '1.1', str_length(SYMBOL) >= 4)   %>% 
  distinct() %>% 
  select(-NATIVITY)
  
  # FOR EVERY SPECIES IN THESE DATASETS, WHICH ARE ALSO IN OUR ATTRIBUTE TABLES
  # WE WILL APPEND OUR OWN FUNCTIONAL DATA TO ENSURE THAT WE HAVE THE 
  # APPROPRIATE MAPPING BETWEEN OUR AND THESE DATA. THE DEFAULTS FROM THESE DATA
  # WILL BE USED FOR REMAINING UNMATCHED RECORDS. 
dp1 <- dominant_plants %>% 
  select(ECO.SITE:UPPER, FUNCTIONAL) %>% 
  left_join(., attributes, by = c('SYMBOL' = 'SYMBOL_USDA')) %>% 
  drop_na(BINOMIAL_NAT) %>% 
  distinct() %>% 
  relocate(any_of(c("PHOTOSYNTHETIC", "RESPROUT", "DURATION", 'GROWTHHABITSUB')), 
           .after = last_col()) %>% 
  unite(col = FUNCTIONAL_FINE, PHOTOSYNTHETIC:GROWTHHABITSUB, 
        sep = "-", na.rm = T, remove = F) %>% 
  select(-PHOTOSYNTHETIC, -RESPROUT) %>% 
  mutate(
    FUNCTIONAL_FINE = str_remove(FUNCTIONAL_FINE, 'CAM-'), 
    FUNCTIONAL_FINE = str_replace(FUNCTIONAL_FINE, 'SUCCULENT|SUBSHRUB', 'SHRUB'),
    FUNCTIONAL_FINE = str_replace(FUNCTIONAL_FINE, 'PERENNIAL-SHRUB', 'SHRUB'), 
    FUNCTIONAL_FINE = str_replace(FUNCTIONAL_FINE, 'PERENNIAL-TREE', 'TREE'), 
    FUNCTIONAL_FINE = str_replace(FUNCTIONAL_FINE, 'GRAMINOID|SEDGE', 'GRASS')
      ) %>% 
  filter(SYMBOL != 'ECTR' & FUNCTIONAL_FINE != 'PERENNIAL-FORB') %>% 
  select(ECO.SITE:UPPER, FUNCTIONAL_FINE, LIFECYLE =  DURATION, FUNCTIONAL = FUNCTIONAL.x) %>% 
  distinct() 
  
dominant_plants <- dominant_plants %>% 
  filter(!SYMBOL %in% unique(dp1$SYMBOL)) %>% 
  bind_rows(., dp1) %>% 
  filter(FUNCTIONAL_FINE != 'GRASS') %>% # not enough info
  mutate(
    FUNCTIONAL_FINE = str_replace(FUNCTIONAL_FINE, 'ANY-FORB|ANNUAL-PERENNIAL-FORB', 'FORB'),
    FUNCTIONAL_FINE = case_when(
      SYMBOL == 'POTEN' ~ 'PERENNIAL-FORB', 
      SYMBOL == 'ARPUL' ~ 'C4-PERENNIAL-GRASS',
      FUNCTIONAL_FINE == 'C3-GRASS' & LIFECYCLE == 'PERENNIAL' ~ 'C3-PERENNIAL-GRASS',
      FUNCTIONAL_FINE == 'C3-GRASS' & LIFECYCLE == 'ANNUAL' ~ 'C3-ANNUAL-GRASS',
      FUNCTIONAL_FINE == 'C4-GRASS' & LIFECYCLE == 'PERENNIAL' ~ 'C4-PERENNIAL-GRASS',
      TRUE ~ FUNCTIONAL_FINE
    )
  )

dominant_plants <- dominant_plants %>% 
  mutate(FUNC_LIFE = case_when(
    FUNCTIONAL_FINE == 'TREE' ~ 'T',
    FUNCTIONAL_FINE == 'PERENNIAL-FORB' ~ 'PF',
    FUNCTIONAL_FINE == 'ANNUAL-FORB' ~ 'AF',
    FUNCTIONAL_FINE == 'C3-ANNUAL-GRASS' ~ 'AG',
    str_detect(FUNCTIONAL_FINE, 'PERENNIAL-GRASS') ~ 'PG',
    str_detect(FUNCTIONAL_FINE, 'SHRUB')  ~  'SH'
  ),
  FUNCTIONAL = if_else(SYMBOL %in% c('ERUM', 'PLIN7', 'PECR5'), 'SHRUB', FUNCTIONAL)) 

# total number of plant species per basic life form groups
benchmarks <- dominant_plants %>% 
  group_by(ECO.SITE, FUNCTIONAL) %>% 
  add_count(name = 'Lifeform_count') %>% 
  group_by(ECO.SITE, FUNCTIONAL_FINE) %>% 
  add_count(name = 'Functional_count') %>% 
  select(ECO.SITE, FUNCTIONAL, FUNC_LIFE, FUNCTIONAL_FINE, Lifeform_count, Functional_count) %>% 
  arrange(ECO.SITE) %>% 
  distinct()

rm(dp1, attributes)
```

```{r add ESD to plots}

esd_class <- read.csv( file.path(praw, f[grep('Tracking', f)]) ) %>% 
  filter(STATUS == 'SAMPLED') %>% 
  mutate(ECO.SITE = if_else(ECO.SITE.MATCHED == F, 'UNKNOWN', ECO.SITE)) %>% 
  select(PLOTID = PLOT.ID, ECO.SITE)

spp_richness <- left_join(spp_richness, esd_class)
lpi <- left_join(lpi, esd_class)

rm(esd_class)
```


```{r Number of Species per Ecological Site Plot}

ord <- c("Annual Forb", "Perennial Forb", "Forb",
         "C3 Annual Grass",  "C3 Perennial Grass", "C4 Perennial Grass", 
         "Non-Resprout Shrub", "Resprout Shrub", "Shrub")

lifeform_fill_pal <- c( '#AE0F86', '#860FAE', '#5d0a79',
                                '#37C864', '#2b982e',  '#134314',
                                  '#D4D7D4', '#7f817f', '#323332')
lifeform_fill_pal <- setNames(lifeform_fill_pal, ord)

benchmarks <- benchmarks %>% 
  ungroup() %>% 
  mutate(plotting_names = stringr::str_to_title(gsub('-', " ", FUNCTIONAL_FINE)) ,
         plotting_names = str_replace(plotting_names, 
                                      'Non Resprout Shrub', 'Non-Resprout Shrub'), 
         plotting_names = factor(plotting_names, levels = ord))

funct_names <- as_labeller(
  c(`FORB` = 'Forb', `GRASS` = 'Grass', `SHRUB` = 'Shrub'))
```

```{r Number of Species per Ecological Site Plot prep, eval = F}

benchmarks %>% 
  filter(FUNCTIONAL_FINE != 'TREE') %>% 
  ggplot() + 
  geom_bar(aes(x = Lifeform_count, fill = plotting_names)) +
  scale_fill_manual('', values = lifeform_fill_pal) +
  facet_wrap(~ FUNCTIONAL,
             labeller = funct_names) +
  theme_bw() +
  theme(legend.position = 'bottom',
        plot.title = element_text(hjust = 0.5),
        legend.spacing.x = unit(1.0, 'cm')) +
  guides(fill = guide_legend(ncol = 3, title.position="top")) +
  labs(x = 'Species per broad Functional Group', y = 'Species per fine Functional',
       title = 'Number of species per functional group across ESD')

ggsave(plot = last_plot(),  device = 'png', height = 10, units = 'cm', 
       path = '../plots/graphics', filename = 'spp_per_ESD.png')
```

In order to determine which, and how many, species are noted to be dominant members of the vegetation at an Ecological Site, all ESD's for which an AIM plot was verified to were checked. All conditions (e.g. all State/Phase combinations which were present) from the reference tables were copied from these manually into spreadsheets, and data were cleaned using R. The USDA symbols utilized in these tables, were verified to match to the look up tables which were created for our project, to ensure that the same species between the two data sources could not be 'missed' due to using different abbreviation codes. Given that only two species (*Pinus edulis*, *Junipers osteosperma*) were ever considered to be true 'Trees' under both systems, this lifeforms was dropped from all analyses to focus on the more variable groups. 

\begin{wrapfigure}{l}{0.5\textwidth}
  \centering
    \includegraphics[width=0.5\textwidth]{../plots/graphics/spp_per_ESD.png}
  \caption{Number of Species Per Functional Groups}
\end{wrapfigure}

In order to be able to make comparisons across functional groups, for each species present in both the ESD's and our plot based data, we ensured that they drew from the same attribute table. This was especially important for situations where sub shrubs are alternatively classified as forbs or shrubs depending on context. After performing this, we recovered both the 'coarse' functional groups used by the AIM team, which correspond largely to the botanical notion of 'lifeforms'. Subsequently, using the developed attribute tables finer functional groups, developed locally, which reflect the propensity of major groups of these lifeforms to respond to various disturbances, such as drought (e.g. C3 and C4 grasses), and fire (non-resprout and re-sprouting shrubs). In instances where more fine scale functional data were not available, such as was the case for sprouting potential of many shrub species, then species were left at this level rather than attempting to determine an appropriate group for them.

The calculation of all, previously mentioned, forms of diversity were not carried out for each individual plot. All plots had $\alpha-diversity$, the number of unique species per plot calculated. However, plots which lacked Ecological Sites, were unable to have $\gamma-diversity$, and subsequently $\beta-diversity$ calculated. Ecological Sites which lacked reference tables were able to have these calculations performed, as they are independent of the Descriptions. Calculations of $\beta-diversity$, wherein only plots belonging to the same Ecological Sites were compared, were performed using the 'vegdist' function from the package 'vegan', with standard defaults and as Sorenson-Dice dissimilarity (@oksanen2022vegan, @sorensen1948method, @dice1945measures). 

$$
\text{Sorenson-Dice Index} = \frac{(2\times\text{No. Species Site 1} \cap \text{Site 2})}{(\text{No. Species Site 1} + \text{No. Species Site 2})}
$$

Wherein the number of species which are shared ("$\cap$") between two sites are multiplied by 2, and divided by the total number of species at both sites.

The data collected by two distinct methods, Line-Point Intercept, and Species Richness, were both compared to the ESD reference data to determine which more closely reflected the benchmark values. After visual exploration of these data displayed in several plots (including Figure 2), it was determined that the Species Richness data more adequately reflected the Benchmarks values and it was used for all subsequent analyses. 

```{r Number of species observed at plots with known ESD}

spp_richness_lifeform <- spp_richness %>% 
  group_by(PLOTID, GROWTHHABITSUB) %>% 
  select(-LIFECYCLE) %>% 
  
  add_count(name = 'Lifeform_count') %>% 
  group_by(PLOTID, FUNCTIONAL_FINE) %>% 
  add_count(name = 'Functional_count')
  
lpi_lifeform <- lpi %>% 
  group_by(PLOTID, GROWTHHABITSUB) %>% 
  select(-LIFECYCLE,  -BINOMIAL_ACK) %>% 
  
  add_count(name = 'Lifeform_count') %>% 
  group_by(PLOTID, FUNCTIONAL_FINE) %>% 
  add_count(name = 'Functional_count')

```


```{r number of plots meeting coarse benchmarks for known ESD}

spp_richness_lifeform_esd <- spp_richness_lifeform %>% 
  filter(ECO.SITE %in% benchmarks$ECO.SITE) %>% 
  distinct(PLOTID, FUNCTIONAL_C = GROWTHHABITSUB, Observed = Lifeform_count, ECO.SITE) %>% 
  left_join(., select(benchmarks, ECO.SITE, FUNCTIONAL_C = FUNCTIONAL, 
                      Expected = Lifeform_count) %>% distinct(), 
            by = c('ECO.SITE' , 'FUNCTIONAL_C')) %>% 
  filter(FUNCTIONAL_C != 'TREE') %>% 
  mutate(METHOD = 'Richness')

lpi_lifeform_esd <- lpi_lifeform %>% 
  ungroup() %>% 
  filter(ECO.SITE %in% benchmarks$ECO.SITE) %>% 
  distinct(PLOTID, FUNCTIONAL_C = GROWTHHABITSUB, Observed = Lifeform_count, ECO.SITE) %>% 
  left_join(., select(benchmarks, ECO.SITE, FUNCTIONAL_C = FUNCTIONAL, 
                      Expected = Lifeform_count) %>% distinct(), 
            by = c('ECO.SITE' , 'FUNCTIONAL_C')) %>% 
  filter(FUNCTIONAL_C != 'TREE')  %>% 
  mutate(METHOD = 'LPI')

lifeform_coarse <- bind_rows(spp_richness_lifeform_esd, lpi_lifeform_esd)

rm(lpi_lifeform_esd)
```

```{r  plot number of plots meeting coarse benchmarks for known ESD, eval = F}

f <- lifeform_coarse %>% 
  filter(FUNCTIONAL_C == 'FORB') %>% 
  group_by(ECO.SITE) %>% 
  mutate(ECO.SITE = fct_reorder(ECO.SITE, Expected, .fun='median')) %>% 
  ggplot(aes(x=reorder(ECO.SITE, Expected), y = Expected))  + 
  geom_point(color = '#AE0F86', size = 4, shape = 18) + 
  
  scale_color_manual(values = c('#e54c09', '#237cd4')) +
  geom_jitter(aes(y = Observed, color = METHOD), width = 0.2, alpha = 0.4) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1, vjust = 1),
        legend.position = "none", 
        panel.background = element_rect(fill='transparent'),
        plot.background = element_rect(fill='transparent', color=NA),
        rect = element_rect(fill = "transparent"),
        plot.margin = unit(c(0.2,0.2,0.2,0.2), "lines"),
        plot.title = element_text(hjust = 0.5)) +
  ylim(0,30) +
  labs(title = 'Forb', y = 'No. Species', x = NULL)

g <- lifeform_coarse %>% 
  filter(FUNCTIONAL_C == 'GRASS') %>% 
  group_by(ECO.SITE) %>% 
  mutate(ECO.SITE = fct_reorder(ECO.SITE, Expected, .fun='median')) %>% 
  ggplot(aes(x=reorder(ECO.SITE, Expected), y = Expected))  + 
  geom_point(color = '#37C864', size = 4, shape = 18) + 
  scale_color_manual(values = c('#e54c09', '#237cd4')) +
  
  geom_jitter(aes(y = Observed, color = METHOD), width = 0.2, alpha = 0.4) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1, vjust = 1),
        legend.position = "none",
        panel.background = element_rect(fill='transparent'),
        rect = element_rect(fill = "transparent"),
        plot.background = element_rect(fill='transparent', color=NA),
        plot.margin = unit(c(0.2,0.2,0.2,0.2), "lines"),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        plot.title = element_text(hjust = 0.5)) +
  ylim(0,30) +
  labs(title = 'Grass', y = NULL, x = NULL)

s <- lifeform_coarse %>% 
  filter(FUNCTIONAL_C == 'SHRUB') %>% 
  group_by(ECO.SITE) %>% 
  mutate(ECO.SITE = fct_reorder(ECO.SITE, Expected, .fun='median')) %>% 
  ggplot(aes(x=reorder(ECO.SITE, Expected), y = Expected))  + 
  geom_point(color = '#7f817f', size = 4, shape = 18) + 
  geom_jitter(aes(y = Observed, color = METHOD), width = 0.2, alpha = 0.4) + 
  scale_color_manual(values = c('#e54c09', '#237cd4')) +
  
  theme_bw() +
  ylim(0,30) +
  theme(
    axis.text.x = element_text(angle = 50, hjust = 1, vjust = 1),
    rect = element_rect(fill = "transparent"),
    panel.background = element_rect(fill='transparent'),
    plot.background = element_rect(fill='transparent', color=NA),
    legend.position = "none",
    plot.margin = unit(c(0.2,0.2,0.2,0.2), "lines"),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    plot.title = element_text(hjust = 0.5)) +
  labs(title = 'Shrub', y = NULL, x = NULL)

legend <- get_legend(
  
  lifeform_coarse %>% 
  ggplot(aes(x= ECO.SITE, y = Expected))  + 
  geom_jitter(aes(y = Observed, color = METHOD), width = 0.2, alpha = 0.4) + 
  scale_color_manual('Method', values = c('#e54c09', '#237cd4')) +
  theme(legend.key=element_blank(),
        legend.background=element_blank())
  
)

plot <- plot_grid(f, g, s, nrow = 1, align='vh', vjust=1, scale = 1)
plot <- plot_grid(plot, legend, ncol = 2, rel_widths = c(1, 0.15))

title <- ggdraw() + draw_label("Species Observed per AIM Plot", size = 16)
x_axis <- ggdraw() + draw_text('Ecological Site', size = 14)

plot_titles <- cowplot::plot_grid(
  title,
  plot, 
  x_axis, 
  ncol = 1, rel_heights = c(0.1, 1, 0.1))

ggsave(plot_titles,  device = 'png', height = 12, units = 'cm', bg = 'transparent',
       path = '../plots/graphics', filename = 'lifeform_esd.png')
  
rm(plot, s, g, f)
```

```{r number of plots meeting fine benchmarks for known ESD}

spp_richness_functional_esd <- spp_richness_lifeform %>% 
  filter(ECO.SITE %in% benchmarks$ECO.SITE) %>% 
  distinct(PLOTID, FUNCTIONAL_FINE, Observed = Lifeform_count, ECO.SITE) %>% 
  left_join(., select(benchmarks, ECO.SITE, FUNCTIONAL_FINE, 
                      Expected = Lifeform_count) %>% distinct(), 
            by = c('ECO.SITE' , 'FUNCTIONAL_FINE')) %>% 
  filter(FUNCTIONAL_FINE != 'TREE') %>% 
  mutate(METHOD = 'Richness')

lpi_functional_esd <- lpi_lifeform %>% 
  filter(ECO.SITE %in% benchmarks$ECO.SITE) %>% 
  distinct(PLOTID, FUNCTIONAL_FINE, Observed = Lifeform_count, ECO.SITE) %>% 
  left_join(., select(benchmarks, ECO.SITE, FUNCTIONAL_FINE, 
                      Expected = Lifeform_count) %>% distinct(), 
            by = c('ECO.SITE' , 'FUNCTIONAL_FINE')) %>% 
  filter(FUNCTIONAL_FINE != 'TREE') %>% 
  mutate(METHOD = 'LPI')

lifeform_fine <- bind_rows(spp_richness_functional_esd, lpi_functional_esd)

rm(lpi_functional_esd, benchmarks)
```

```{r fine lifeform species diverity plot, eval = F}

## col 1,  row 1 & 2  ##    

pg3 <- lifeform_fine %>% 
  filter(FUNCTIONAL_FINE == 'C3-PERENNIAL-GRASS') %>% 
  group_by(ECO.SITE) %>% 
  mutate(ECO.SITE = fct_reorder(ECO.SITE, Expected, .fun='median')) %>% 
  ggplot(aes(x=reorder(ECO.SITE, Expected), y = Expected))  + 
  geom_point(color = '#2b982e', size = 4, shape = 18)+ 
  geom_jitter(aes(y = Observed, color = METHOD), width = 0.2, alpha = 0.4) + 
  scale_color_manual(values = c('#e54c09', '#237cd4'))  + 
  
  theme_bw() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1, vjust = 1),
        legend.position = "none",
        plot.title = element_text(hjust = 0.5)) +
  labs(title = 'Perennial C3 Grass', y = NULL, x = NULL)

pg4 <- lifeform_fine %>% 
  filter(FUNCTIONAL_FINE == 'C4-PERENNIAL-GRASS') %>% 
  group_by(ECO.SITE) %>% 
  mutate(ECO.SITE = fct_reorder(ECO.SITE, Expected, .fun='median')) %>% 
  ggplot(aes(x=reorder(ECO.SITE, Expected), y = Expected))  + 
  geom_point(color = '#134314', size = 4, shape = 18) + 
  geom_jitter(aes(y = Observed, color = METHOD), width = 0.2, alpha = 0.4) + 
  scale_color_manual(values = c('#e54c09', '#237cd4')) + 
  
  theme_bw() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1, vjust = 1),
        legend.position = "none",
        plot.title = element_text(hjust = 0.5)) +
  labs(title = 'Perennial C4 Grass', y = NULL, x = NULL)

## col 2, rows 1 & 2  ##

sh_s <- lifeform_fine %>% 
  filter(FUNCTIONAL_FINE == 'RESPROUT-SHRUB') %>% 
  group_by(ECO.SITE) %>% 
  mutate(ECO.SITE = fct_reorder(ECO.SITE, Expected, .fun='median')) %>% 
  ggplot(aes(x=reorder(ECO.SITE, Expected), y = Expected))  + 
  geom_point(color = '#7f817f', size = 4, shape = 18) +
  geom_jitter(aes(y = Observed, color = METHOD), width = 0.2, alpha = 0.4) + 
  scale_color_manual(values = c('#e54c09', '#237cd4'))  + 
  
  theme_bw() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1, vjust = 1),
        legend.position = "none",
        plot.title = element_text(hjust = 0.5)) +
  labs(title = 'Resprouting Shrubs', y = NULL, x = NULL)

sh_ns  <- lifeform_fine %>% 
  filter(FUNCTIONAL_FINE == 'NON-RESPROUT-SHRUB') %>% 
  group_by(ECO.SITE) %>% 
  mutate(ECO.SITE = fct_reorder(ECO.SITE, Expected, .fun='median')) %>% 
  ggplot(aes(x=reorder(ECO.SITE, Expected), y = Expected))  + 
  geom_point(color = '#323332', size = 4, shape = 18) + 
  geom_jitter(aes(y = Observed, color = METHOD), width = 0.2, alpha = 0.4) + 
  scale_color_manual(values = c('#e54c09', '#237cd4')) +
  
  theme_bw() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1, vjust = 1),
        legend.position = "none",
        plot.title = element_text(hjust = 0.5)) +
  labs(title = 'Non-Resprouting Shrubs', y = NULL, x = NULL)

## col 3, rows 1 & 2 ##

af <- lifeform_fine %>% 
  filter(FUNCTIONAL_FINE == 'ANNUAL-FORB') %>% 
  group_by(ECO.SITE) %>% 
  mutate(ECO.SITE = fct_reorder(ECO.SITE, Expected, .fun='median')) %>% 
  ggplot(aes(x=reorder(ECO.SITE, Expected), y = Expected))  + 
  geom_point(color = '#AE0F86', size = 4, shape = 18)+ 
  geom_jitter(aes(y = Observed, color = METHOD), width = 0.2, alpha = 0.4) + 
  scale_color_manual(values = c('#e54c09', '#237cd4')) + 
  
  theme_bw() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1, vjust = 1),
        legend.position = "none",
        plot.title = element_text(hjust = 0.5)) +
  labs(title = 'Annual Forbs', y = NULL, x = NULL)

pf <- lifeform_fine %>% 
  filter(FUNCTIONAL_FINE == 'PERENNIAL-FORB') %>% 
  group_by(ECO.SITE) %>% 
  mutate(ECO.SITE = fct_reorder(ECO.SITE, Expected, .fun='median')) %>% 
  ggplot(aes(x=reorder(ECO.SITE, Expected), y = Expected))  + 
  geom_point(color = '#860FAE', size = 4, shape = 18)+ 
  geom_jitter(aes(y = Observed, color = METHOD), width = 0.2, alpha = 0.4) + 
  scale_color_manual(values = c('#e54c09', '#237cd4')) + 
  
  theme_bw() +
  theme(axis.text.x = element_text(angle = 50, hjust = 1, vjust = 1),
        legend.position = "none",
        plot.title = element_text(hjust = 0.5)) +
  labs(title = 'Perennial Forbs', y = NULL, x = NULL)

top <- plot_grid(af, pg3, sh_ns, nrow = 1, align='vh', vjust=1, scale = 1)
bottom <- plot_grid(pf, pg4, sh_s, nrow = 1, align='vh', vjust=1, scale = 1)
plot_grid(top, bottom, ncol = 1)

ggsave(plot = last_plot(),  device = 'png', height = 15, units = 'cm', 
       path = '../plots/graphics', filename = 'functional_esd.png')
  
rm(af, pf, pg3, pg4, sh_s, sh_ns)
```


```{r Proportion of Species in Production Table Per ESD by Functional}

prop_lf_c <- spp_richness_lifeform_esd %>% 
  rowwise() %>% 
  mutate(Expected = replace_na(Expected, 0),
         Proportion = if_else(Observed < Expected, Observed/Expected, 1)) %>% 
  group_by(ECO.SITE, FUNCTIONAL_C) %>% 
  reframe(qs = quantile(Proportion, c(0.25, 0.75)), prob = c(0.25, 0.75), 
          n = n(), med = median(Proportion))

prop_lf_f <- spp_richness_functional_esd %>% 
  rowwise() %>% 
  mutate(Expected = replace_na(Expected, 0),
         Proportion = if_else(Observed < Expected, Observed/Expected, 1)) %>% 
  group_by(ECO.SITE, FUNCTIONAL_FINE) %>% 
  reframe(qs = quantile(Proportion, c(0.25, 0.75)), prob = c(0.25, 0.75), 
          n = n(), med = median(Proportion))

rm(spp_richness_functional_esd, spp_richness_lifeform_esd, lpi_lifeform, lpi_lifeform_esd)
```

```{r gamma diversity by ESD}

# library(vegan) # are both required. 
# library(reshape2)

alpha_diver <- spp_richness %>% 
  group_by(PLOTID) %>% 
  count(name = 'Alpha') 

gamma_diver <- spp_richness %>% 
  group_by(ECO.SITE) %>% 
  distinct(SYMBOL) %>% 
  count(name = 'Gamma') %>% 
  filter(ECO.SITE != 'UNKNOWN')

# beta diversity for many sites, Sorenson-Dice dissimilarity

beta <- spp_richness %>% 
  st_drop_geometry() %>% 
  filter(ECO.SITE != 'UNKNOWN') %>% 
  group_by(ECO.SITE) %>% 
  mutate(Replicates = n_distinct(PLOTID),
         PRESENCE = 1) %>% 
  filter(Replicates >= 2) %>% 
  ungroup() %>% 
  select(PLOTID, SYMBOL, ECO.SITE, PRESENCE) %>% 
  split(.$ECO.SITE) %>% 
  map(. %>%     # arrange data into a binary matrix format required for the
        select(-ECO.SITE) %>%  # function
        pivot_wider(
          names_from = 'SYMBOL',
          values_from = PRESENCE, 
          values_fill = 0, 
          values_fn = function(x) 1) %>% 
        column_to_rownames('PLOTID') 
      )  

beta_est <- lapply(beta, vegdistR) %>% 
  bind_rows(., .id = 'ECO.SITE')

diversity <- beta_est %>% 
  pivot_longer(Plot1:Plot2, values_to = 'PLOTID') %>% 
  group_by(PLOTID) %>% 
  mutate(MeanBeta = mean(Sorenson)) %>% 
  select(-name, -Sorenson) %>% 
  distinct(PLOTID, .keep_all = T) %>% 
  left_join(., gamma_diver) %>% 
  right_join(., alpha_diver) %>% 
  st_as_sf() %>% 
  group_by(ECO.SITE) %>% 
  mutate(N_plots = n(), .before = 'geometry')

rm(beta, vegdistR, gamma_diver, alpha_diver, beta_est)
```


```{r what is the turnover in both core functional and species richness between plots of same Ecological Site, eval = F}
library(terra)

p2carto <- '../../UFO_cartography'
ext <- vect('../../UFO_cartography/UFO_mask/mask/mask.shp') 

elev_90 <- rast(file.path(p2carto, list.files(p2carto, recursive = T, pattern = 'DEM.*tif')))
div_v <- vect(diversity) |> project(crs(elev_90))

elevation <- extract(elev_90, div_v)
names(elevation) <- c('id', 'elev')

diversity1 <- bind_cols(diversity, 'elev' =  elevation[,2])

summary( lm(Alpha ~ elev, data = diversity1) )

ggplot() +
  geom_point(data = diversity1, aes(x = N_plots, y = Gamma)) +
  geom_smooth()

rm(p2carto, ext, elev_90, div_v, diversity1)
```

In order to determine whether enough plots have been sampled to reliably estimate the number of species which are present per Ecological Site ($\gamma-diversity$), rarefaction and extrapolation curves from the iNEXT package were used with confidence intervals of 0.8 (@Hsieh2022iNEXT, @Chao2014iNEXT).

```{r Rarefaction by Ecological site - how many species are extrapolated}

rarefaction <- spp_richness %>% 
  st_drop_geometry() %>% 
  filter(ECO.SITE != 'UNKNOWN') %>% 
  group_by(ECO.SITE) %>% 
  mutate(Replicates = n_distinct(PLOTID),
         PRESENCE = 1) %>% 
  filter(Replicates >= 2) %>% 
  ungroup() %>% 
  select(PLOTID, SYMBOL, ECO.SITE, PRESENCE) %>% 
  split(.$ECO.SITE) %>% 
  map(. %>%     # arrange data into a binary matrix format required for the
        select(-ECO.SITE) %>%  # function
        pivot_wider(
          names_from = 'SYMBOL',
          values_from = PRESENCE, 
          values_fill = 0, 
          values_fn = function(x) 1) %>% 
        column_to_rownames('PLOTID') %>% 
        t()
      )  

rarefaction <- i_report(rarefaction, conf.int = 0.8, k = 25)
```


```{r Clear methods environment }
rm(f, fp, funct_names, lifeform_fill_pal, ord, ppro, praw)
```

# Discussion

\begin{wrapfigure}{l}{0.6\textwidth}
  \centering
    \includegraphics[width=0.5\textwidth]{../plots/graphics/lifeform_esd.png}
  \caption{Number of Species Per Coarse Functional Groups}
\end{wrapfigure}

```{r}
m <- diversity %>% 
  distinct(ECO.SITE, .keep_all = T) %>% 
  st_drop_geometry() %>% 
  drop_na() %>% 
  ungroup()
```

Species richness data contained a total of `r no_records_full` records, which after removal of unidentified material left `r sum(diversity$Alpha)` species records at `r nrow(diversity)` plots which were verified to Ecological Sites. `r nrow(m)` Ecological Sites had more than two AIM plots located in them, and had $\beta-diversity$ calculated for all plots `r sum(m$N_plots)` located in them.

While the species within any one reference table in an Ecological Site Descriptions are noted to represent the dominant species across a gradient which stratifies the Ecological Site, such as elevation, many of our plots contained an adequate number of these species per functional group. In fact, when considering the richness data, which we suspect is a more apt dataset for comparing to the ESD taxa, many of our plots had the number of species expected in the reference table, for both the Forb and Shrub functional groups. For Shrubs it appears that the Ecological Sites which are meant to have the greatest number of shrub species do not reach these goals; and that perhaps only for these sites is the species representing the gradients of the Ecological Site utilized. Accordingly, it is unlikely that these tables represent variables collected in a consistent manner; regardless they offer clear insights. The number of grass species, in only a handful of plots, meets or exceeds the numbers of grass species noted present in reference conditions in all ESD's. Anecdotally, grass species which appear to be missing from expanses of the field office include two species of *Sporobolus* (*S. airoides*, and *S. cryptandrus*), *Koeleria macrantha*, with noted declines in *Hesperostipa comata* in areas as well. 



~ ALPHA DIVERSITY HERE ~

```{r alpha diversity results}

alphaD <- diversity %>% 
  ungroup() %>% 
  slice_max(Alpha, n = 25) %>% 
  st_drop_geometry() %>% 
  group_by(ECO.SITE) %>% 
  add_count()

ad <- alphaD %>% 
  ungroup() %>% 
  drop_na(ECO.SITE) %>% 
  distinct(ECO.SITE, .keep_all = T) %>% 
  slice_max(n, n = 5) %>% 
  mutate(
    PlotsTop = paste0(ECO.SITE, ' (plots = ', n, ')') ) %>% 
  pull(PlotsTop)
  
```

The number of identified plant species varied widely by plot, from `r min(diversity$Alpha)` - `r max(diversity$Alpha)` (Med = `r median(diversity$Alpha)`). 1) The `r nrow(alphaD)` plots with the most species in them were in `r nrow(unique(alphaD[!is.na(alphaD$ECO.SITE), 'ECO.SITE']))` Ecological Sites, this subset of plots had a minimum of `r min(alphaD$Alpha)` species per plot. Of the top `r nrow(alphaD)` plot subset `r nrow(ad)` the Ecological Sites with the highest observed $\alpha-diversity$ were: `r toString(ad)`

3) XXX %, % , % of all Ecological Sites had fewer species of forbs, grasses, and shrubs observed than expected. 4) XXX % of all plots had fewer Annual Forbs, Perennial Forbs, or unclassified forbs than expected. XXX % of all plots had fewer Annual C3 Grasses, Perennial C3 Grasses, C4 Perennial Grasses than expected. XXX $ of all plots had fewer Non-Resprouting Shrubs, Resprouting, and unclassified shrubs than expected. 

~ BETA DIVERSITY HERE ~
```{r beta diversity results}

betaD <- diversity %>% 
  ungroup() %>% 
  slice_max(MeanBeta, n = 25) %>% 
  st_drop_geometry() %>% 
  group_by(ECO.SITE) %>% 
  add_count()



bd <- betaD %>% 
  ungroup() %>% 
  drop_na(ECO.SITE) %>% 
  distinct(ECO.SITE, .keep_all = T) %>% 
  slice_max(n, n = 5) %>% 
  mutate(
    PlotsTop = paste0(ECO.SITE, ' (', n, ')') ) %>% 
  pull(PlotsTop)

rm(ad, alphaD)
```

The mean beta diversity, that is the dissimilarity between plots varied from `r round(min(m$MeanBeta),2)` - `r round(max(m$MeanBeta),2)`, with lower values indicating greater differences between plots; and where the Ecological site with the most dissimilar plots, on average, contained just over half (`r round(min(m$MeanBeta),2)`) of the same species as each other. The range of Sorenson-Dice values for the `r nrow(betaD)` most dissimilar plots were from `r min(betaD$MeanBeta)` - `r max(betaD$MeanBeta)`.  The `r nrow(betaD)` most dissimilar plots were in `r nrow(unique(betaD[!is.na(betaD$ECO.SITE), 'ECO.SITE']))`, with the following Ecological Sites having the most plots in them: `r toString(bd)`
 
 
 . 3) The beta-diversity of the major functional groups varied from MIN to MAX, for forbs, grasses, and shrubs respectively. This indicating that turnover in FORBS is much greater than for the other two groups. The turnover in species composition for the minor functional groups ... 

Estimates of $\beta-diversity$ appear less influenced by the number of plots sampled and $\gamma-diversity$, or surprisingly elevation. The discord between elevation is surprising given that $\beta-diversity$ was initially conceived to largely operate along elevation gradients, and oftentimes does (CITE). 


~ GAMMA DIVERSITY HERE ~

```{r Gamma diversity results}
NoSppGamma <- rarefaction %>% 
  slice_max(SpeciesObserved, n = 5) %>% 
  mutate(
    NoSppObs = paste0(Eco.Site, ' (spp. = ', SpeciesObserved, ')'),
    round(across(.cols = SpeciesEstimator:UCL, ~ round(.x, 0))), 
    NoSppPred = paste0(Eco.Site, ' (spp pred. = ', SpeciesEstimator, ', LCL = ', LCL,
                       ', UCL = ',  UCL, ')' )
    ) 
```

Ecological Sites varied in the number of species which were observed in them. However, the accuracy of the current estimates of $\gamma-diversity$ are limited by the number of replicates per Ecological Site. The estimated Sample coverage per Ecological Site ranged from `r min(rarefaction$SampleCoverage)` to `r max(rarefaction$SampleCoverage)` with a median of `r median(rarefaction$SampleCoverage)`, 80% confidence interval; in general it appears only 3-4 plots are required to gain Sample coverage estimates about 50%. In other words, we do not yet have enough replicates per Ecological Site to make definitive statements regarding the species richness of each site, but clear trends have emerged. The Ecological Sites with the greatest number of species are `r toString(NoSppGamma$NoSppObs)`. The predicted estimates of the number of species which are to be expected, in the UFO, at these sites with further sampling are: `r toString(NoSppGamma$NoSppPred)` (80% confidence intervals).

**Most of these sites are likely to have the highest number of estimated  species, due to the size of their geographic expanses, fostering high rates of $\beta-diversity$ rather than high $\alpha-diversity$.**

The Ecological Sites with the greatest functional diversity have XXX XXXX XXX species of forbs, grasses, and shrubs respectively. The five ecological sites with the greatest functional diversity of forbs, have XXX XXX XXX annual forbs, perennial forbs, and unclassified forbs.  
The five ecological sites with the greatest functional diversity of grasses, have XXX XXX XXX C3 annual grass, C3 perennial grass, and C4 perennial grasses.
The five ecological Sites with the greatest functional diversity of shrubs have XXX XXX XXX Non-Resprouting Shrubs, Resprouting, and unclassified shrubs. 

Currently, two Ecological sites with `r slice_max(m, N_plots, n = 2, with_ties = F) |> pull(N_plots) |> toString()` sampled plots contain `r slice_max(m, N_plots, n = 2, with_ties = F) |> pull(Gamma) |> toString()` unique identified species. Given the total number of plants observed on AIM plots (Section X.X), this constitute environments holding nearly a 1/3 of this $\alpha-diversity$ themselves. 

```{r clear writing variables from environment}
rm(lifeform_esd, no_records_full, diversity, dominant_plants,
   lifeform_coarse, lifeform_fine, lpi, prop_lf_c, prop_lf_f, spp_richness,
   spp_richness_lifeform, spp_richness_clean)

```

\newpage

# References

